### 2020-02-20 Writing/disparity_review.tex
updated references

\hypertarget{introduction}{%
\section{1. Introduction}\label{introduction}}

Clades of organisms are characterised by variation in both numbers of species and range of phenotypes through time.
At the extremes, clades may be exceptionally species-rich and phenotypically diverse (e.g. cichlids and molluscs), species-rich but phenotypically conservative (e.g. bacteria and nematodes), species-poor but phenotypically diverse (e.g. Afrotheria), or depauperate in both species and phenotypic diversity (e.g. lungfish).
These phenomena suggest that taxonomic and phenotypic diversity are not inextricably linked, raising important questions about how phenotypic diversity arises, such as:
How does diversity evolve?
Are some morphologies more common than others?
Can anatomy evolve in all ``directions'' or are some anatomies impossible to achieve? (Mike Foote 1997, 1995)
What role does ecology play in structuring morphological diversity?
Analyses of species diversity have a venerable history, but those of phenotypic diversity (hereafter \emph{disparity}) are a comparatively more recent phenomenon. Originally defined as ``multidimensional morphological dissimilarity at a macroevolutionary scale'' (Runnegar 1987, Gould \href{https://paperpile.com/c/sTGYvp/Uns3}{@@@}),
% Runnegar, Bruce "Rates and modes of evolution in the Mollusca." Rates of evolution. Allen and Unwin, London (1987): 39-60.
 this concept of disparity emerged from attempts by palaeobiologists to characterise the evolutionary origin of animal bodyplans and from attempts by comparative developmental biologists to provide causal explanations for their emergence.
However, analyses of ``disparity'' have since expanded into comparative biology as a means of capturing the effect of intrinsic and extrinsic causal agents in morphological evolution.
Typically, methods to capture disparity are based on multidimensional spaces where each dimension represents an aspect of morphological variation (a trait) and biological observations (taxa) can be placed in this space based on their trait values.
Such multidimensional spaces (or morphospaces) can then be used to tackle a diverse array of questions that can be grouped into four main (non-mutually exclusive) classes (Fig. 1):

(1) \emph{Descriptive disparity.} The pioneering studies of disparity characterised the shapes of organisms and how they differed among groups \href{https://paperpile.com/c/sTGYvp/fTJ3+eZ3F}{(Foote 1995; Wills, Briggs, and Fortey 1994)}.
These studies consist of describing multidimensional patterns in the diversity of morphological traits, addressing questions such as: why are some morphological trait combinations more common than others and what are the biological (or mathematical) properties of the resulting morphospace? \href{https://paperpile.com/c/sTGYvp/fTJ3+I0Ic+QVvv}{(Foote 1995; Raup 1961; Gerber 2017)}.

(2) \emph{Disparity-through-time.} This approach investigates how the morphologies of organisms have changed on a temporal scale, focussing on the disparity of taxa in particular time intervals or slices.
This approach has been used widely in palaeobiology to answer a range of macroevolutionary questions, such as: how does disparity accumulate over the history of a clade \href{https://paperpile.com/c/sTGYvp/ekU4+s33b}{(Guillerme and Cooper 2018; Wright 2017)}, or how does disparity change leading up to and across mass extinction events \href{https://paperpile.com/c/sTGYvp/EETc}{(Friedman 2010)}?

(3) \emph{Disparity and taxonomic diversity.} Morphological disparity provides another perspective on biodiversity; high morphological disparity represents a high diversity of morphologies (i.e. shapes or body plans) and is, presumably, associated with high levels of ecological and functional diversity (though see section 4 below).
This makes disparity an informative complement to diversity measures based on species richness alone. Indeed, most studies that have investigated disparity and taxonomic diversity support an effective decoupling \href{https://paperpile.com/c/sTGYvp/2tbJ+geAO+hea5+aVVj}{(e.g., Fortey, Briggs, and Wills 1996; Ruta et al. 2013; M. J. Hopkins 2013; Moyne and Neige 2007)}.
This approach has been used to investigate whether some groups are more successful than others in their exploration of new evolutionary strategies \href{https://paperpile.com/c/sTGYvp/dJHu}{(Losos 2011)}.

(4) \emph{Disparity as a proxy for ecology.} The disparity of a group can be used as a proxy for either the functional role it plays within an ecosystem or its ecological niche.
This approach assumes that groups with high disparity are also likely to be functionally and ecologically diverse, and that groups found in similar regions of shape space will have similar functional and ecological roles \href{https://paperpile.com/c/sTGYvp/EETc+tSIy+qjj9}{(Friedman 2010; Pierce, Angielczyk, and Rayfield 2008; P. S. L. Anderson et al. 2011)}.
The links between form and function, however, are not always clear cut.
Traits can be linked to multiple functions and multiple functions can be linked to a single trait \href{https://paperpile.com/c/sTGYvp/Ejzr}{(Wainwright et al. 2005)}.
This approach has been used to investigate hypotheses of competitive replacement (Stephen L. Brusatte et al. 2008) and changes in ecosystem function during and after mass extinctions \href{https://paperpile.com/c/sTGYvp/EETc}{(Friedman 2010)}.
It is particularly common in palaeobiology where it is not possible to directly observe the ecological or functional characteristics of extinct species \href{https://paperpile.com/c/sTGYvp/Ejzr}{(Wainwright et al.} 2005\href{https://paperpile.com/c/sTGYvp/Ejzr}{)}.

(5) Disparity and evo-devo. Morphological disparity blalbalbal
% AG: Although intrinsic constraints are mentioned earlier, it doesn’t enter into any of these topics explicitly, but it is certainly one of the most interesting things one can use disparity for – to bridge development and evolution, looking at developmental biases/constraints. This also includes integration and modularity, which are also central to the topic of disparity and evolvability. Although you touch on these in places, I think it's really important to incorporate it into one of these topics, as there probably isn't room to give it one of its own (although i would argue it warrants equal footing with ecology, given that these are the processes that generate variation in the first place). There is a massive literature on using disparity to test hypotheses of developmental constraints, and also a huge one on disparity and integration/modularity. Let me know if you want me to send you refs.

Fundamental insights into evolutionary biology have been elicited from these five types of analysis.
One of the most important insights is the discovery that morphological disparity is often greatest early in the evolutionary history of clades \href{https://paperpile.com/c/sTGYvp/xxh5+yqPw}{(Hughes, Gerber, and Wills 2013; Foote 1997;} \href{https://paperpile.com/c/sTGYvp/EPJ2}{Erwin 2007)}, indicating that the capacity for evolutionary innovation wanes with clade age, which some have argued reflects the evolutionary assembly of gene regulatory networks that constrain later fundamental change \href{https://paperpile.com/c/sTGYvp/xxh5+Z6l6}{(Hughes, Gerber, and Wills 2013; Erwin 2011)}.
However, this example highlights one of the most challenging problems confronting researchers who are attempting, increasingly, to obtain general insights from multiple independent studies: can the insights gained from studies using the diversity of methods, approaches and data types employed be considered equivalent?

@@

\emph{Figure disparities {[}figure\_disparities.pdf{]}: The four main types of disparity analysis. \textbf{Descriptive disparity} focuses on describing the features of morphospace occupation; \textbf{disparity-through-time} investigates the evolution of the morphospace through time including the effect of extinction events; \textbf{disparity and taxonomic diversity} compares different measures of biodiversity ; \textbf{disparity as a proxy for ecology} uses disparity as a proxy for the ecological or functional role of a group.
These categories are not independent and many disparity studies will cover more than one.}

@@

\hypertarget{data-and-disparity}{%
\section{2. Data and disparity }\label{data-and-disparity}}

Disparity analyses are based on traits, but traits can be characterised in a number of ways:
% MH: We do not mention model-based descriptors at all, or the spaces defined by them (e.g. Raup's shell coiling models). I'm not convinced that we must, but it may come across as an notable omission to reviewers
% SG: I agree. There's for instance Saunders et al. (2004) disparity study of Paleozoic ammonoids using Raup's shell coiling parameters
(1) discrete ``cladistic'' characters, e.g. coding the absence or presence of features or a discrete characteristic of a trait \href{https://paperpile.com/c/sTGYvp/PbSx}{(Close et al. 2015)};
(2) continuous measurements of features (e.g., lengths) \href{https://paperpile.com/c/sTGYvp/qjj9}{(P. S. L. Anderson et al. 2011)}; or
(3) more mathematical descriptors from geometric morphometric landmark data (e.g. Procrustes coordinates) \href{https://paperpile.com/c/sTGYvp/RjqE}{(Cooney et al. 2017)}, and Fourier coefficients \href{https://paperpile.com/c/sTGYvp/2Neu}{(Foote 1989;} \href{https://paperpile.com/c/sTGYvp/ZEDR}{Spriggs et al. 2018)}) (Fig. 2).
None of these approaches are superior, but they may be more or less well-suited to characterising traits under comparison and to the question being asked of those traits (Hetherington et al 2015; Hopkins 2017).

@@

The points above assume that researchers are collecting their own data for disparity analyses, but often this is not the case.
Discrete character data are commonly recycled from phylogenetic studies (e.g.Foote 1989\href{https://paperpile.com/c/sTGYvp/2Neu+0y4V}{; Deline et al. 2018}).
% SG: Foote (1989) is a study of trilobite disparity from Fourier analysis of cranidia. I don't think it has anything to do with discrete data or phylogenetic studies
% MH: Also the Foote papers that used discrete data weren't taken from phylogenetic studies but collected for the specific purpose of the project -- I think this might be true for the Deline paper as well. But there are many studies out there that used matrices initially intended for phylogenetics. I can look for some if needed.
This is an efficient approach to character sampling, but it may artifactually increase disparity between phylogenetically distinct groups because phylogenetic characters are often collected to discriminate among groups \href{https://paperpile.com/c/sTGYvp/fTJ3}{(Foote 1995)}.
This needs to be considered when interpreting results, especially as synapomorphies will naturally lead to an apparent shift or increase in disparity when new clades appear.
Furthermore, many datasets are limited to subsets of anatomy that are at least implicit samples of overall anatomy, but explicit tests of this assumption have shown that different aspects of morphology can exhibit different patterns of disparity \href{https://paperpile.com/c/sTGYvp/xLdm}{(Melanie J. Hopkins 2017)}.
This effect of anatomical part on disparity patterns can be especially challenging when working on datasets where the available data has non-random missing anatomical parts, such as the absence of soft tissue in the fossil record \href{https://paperpile.com/c/sTGYvp/0y4V}{(Deline et al. 2018)}.

>>>>

\section{Introduction}

Clades of organisms are characterised by variation in both numbers of species and range of phenotypes through time.
At the extremes, clades may be exceptionally species-rich and phenotypically diverse (e.g. cichlids and molluscs), species-rich but phenotypically conservative (e.g. bacteria and nematodes), species-poor but phenotypically diverse (e.g. Afrotheria), or depauperate in both species and phenotypic diversity (e.g. lungfish).
These phenomena suggest that taxonomic and phenotypic diversity are not inextricably linked, raising important questions about how phenotypic diversity arises, such as:
How does diversity evolve?
Are some morphologies more common than others?
Can anatomy evolve in all ``directions'' or are some anatomies impossible to achieve? (Mike Foote 1997, 1995)
What role does ecology play in structuring morphological diversity?
Analyses of species diversity have a venerable history, but those of phenotypic diversity (hereafter \emph{disparity}) are a comparatively more recent phenomenon. Originally defined as ``multidimensional morphological dissimilarity at a macroevolutionary scale'' \citep{runnegar1987rates,Gould1991-nh} this concept of disparity emerged from attempts by palaeobiologists to characterise the evolutionary origin of animal bodyplans and from attempts by comparative developmental biologists to provide causal explanations for their emergence.
However, analyses of ``disparity'' have since expanded into comparative biology as a means of capturing the effect of intrinsic and extrinsic causal agents in morphological evolution.
Typically, methods to capture disparity are based on multidimensional spaces where each dimension represents an aspect of morphological variation (a trait) and biological observations (taxa) can be placed in this space based on their trait values.
Such multidimensional spaces (or morphospaces) can then be used to tackle a diverse array of questions that can be grouped into four main (non-mutually exclusive) classes (Fig. 1):

\begin{enumerate}

	\item \textbf{Descriptive disparity} The pioneering studies of disparity characterised the shapes of organisms and how they differed among groups \citep{Foote1995-do, Briggs1992-pd}.
	These studies consist of describing multidimensional patterns in the diversity of morphological traits, addressing questions such as: why are some morphological trait combinations more common than others and what are the biological (or mathematical) properties of the resulting morphospace? \citep{Foote1995-do, Raup1961-vx, Gerber2017-xi}.

	\item \textbf{Disparity-through-time} This approach investigates how the morphologies of organisms have changed on a temporal scale, focussing on the disparity of taxa in particular time intervals or slices.
	This approach has been used widely in palaeobiology to answer a range of macroevolutionary questions, such as: how does disparity accumulate over the history of a clade \citep{Guillerme2018-uj, Wright2017-jo}, or how does disparity change leading up to and across mass extinction events \citep{Friedman2010-ve}?

	\item \textbf{Disparity and taxonomic diversity} Morphological disparity provides another perspective on biodiversity; high morphological disparity represents a high diversity of morphologies (i.e. shapes or body plans) and is, presumably, associated with high levels of ecological and functional diversity (though see section 4 below).
	This makes disparity an informative complement to diversity measures based on species richness alone. Indeed, most studies that have investigated disparity and taxonomic diversity support an effective decoupling 
	\citep[][e.g.]{Fortey1996-kt, Moyne2007-jm, Ruta2013-iy, Hopkins2013-xt}.
	This approach has been used to investigate whether some groups are more successful than others in their exploration of new evolutionary strategies \citep{Losos2011-fq}.

	\item \textbf{Disparity as a proxy for ecology} The disparity of a group can be used as a proxy for either the functional role it plays within an ecosystem or its ecological niche.
	This approach assumes that groups with high disparity are also likely to be functionally and ecologically diverse, and that groups found in similar regions of shape space will have similar functional and ecological roles
	\citep{Friedman2010-ve, Pierce2008-yr, Anderson2013-zt}.
	The links between form and function, however, are not always clear cut.
	Traits can be linked to multiple functions and multiple functions can be linked to a single trait \citep{Wainwright2005-or}.
	This approach has been used to investigate hypotheses of competitive replacement \citep{Brusatte2008-vx} and changes in ecosystem function during and after mass extinctions \citep{Friedman2010-ve}.
	It is particularly common in palaeobiology where it is not possible to directly observe the ecological or functional characteristics of extinct species \citep{Wainwright2005-or}.

	\item \textbf{Disparity and evo-devo} Morphological disparity blalbalbal
% AG: Although intrinsic constraints are mentioned earlier, it doesn’t enter into any of these topics explicitly, but it is certainly one of the most interesting things one can use disparity for – to bridge development and evolution, looking at developmental biases/constraints. This also includes integration and modularity, which are also central to the topic of disparity and evolvability. Although you touch on these in places, I think it's really important to incorporate it into one of these topics, as there probably isn't room to give it one of its own (although i would argue it warrants equal footing with ecology, given that these are the processes that generate variation in the first place). There is a massive literature on using disparity to test hypotheses of developmental constraints, and also a huge one on disparity and integration/modularity. Let me know if you want me to send you refs.

\end{enumerate}

Fundamental insights into evolutionary biology have been elicited from these five types of analysis.
One of the most important insights is the discovery that morphological disparity is often greatest early in the evolutionary history of clades \citep{Foote1997-nl, Erwin2007-jz, Hughes2013-td}, indicating that the capacity for evolutionary innovation wanes with clade age, which some have argued reflects the evolutionary assembly of gene regulatory networks that constrain later fundamental change  \citep{Erwin2007-jz, Hughes2013-td}.
However, this example highlights one of the most challenging problems confronting researchers who are attempting, increasingly, to obtain general insights from multiple independent studies: can the insights gained from studies using the diversity of methods, approaches and data types employed be considered equivalent?

@@

\textit{Figure disparities {[}figure\_disparities.pdf{]}: The four main types of disparity analysis. \textbf{Descriptive disparity} focuses on describing the features of morphospace occupation; \textbf{disparity-through-time} investigates the evolution of the morphospace through time including the effect of extinction events; \textbf{disparity and taxonomic diversity} compares different measures of biodiversity ; \textbf{disparity as a proxy for ecology} uses disparity as a proxy for the ecological or functional role of a group.
These categories are not independent and many disparity studies will cover more than one.}

@@

\section{Data and disparity}

Disparity analyses are based on traits, but traits can be characterised in a number of ways:
% MH: We do not mention model-based descriptors at all, or the spaces defined by them (e.g. Raup's shell coiling models). I'm not convinced that we must, but it may come across as an notable omission to reviewers
% SG: I agree. There's for instance Saunders et al. (2004) disparity study of Paleozoic ammonoids using Raup's shell coiling parameters
	1) discrete ``cladistic'' characters, e.g. coding the absence or presence of features or a discrete characteristic of a trait \citep[][e.g]{Close2015-qi};
	2) continuous measurements of features (e.g., lengths) \citep[][e.g]{Anderson2001-qb}; or
	3) more mathematical descriptors from geometric morphometric landmark data (e.g. Procrustes coordinates) \citep[][e.g]{Cooney2017-ly}, and Fourier coefficients \citep[][e.g]{Foote1995-do, Spriggs2018-nu} (Fig. 2).
None of these approaches are superior, but they may be more or less well-suited to characterising traits under comparison and to the question being asked of those traits \citep{hetherington2015cladistic,Hopkins2017-cf}.

@@

The points above assume that researchers are collecting their own data for disparity analyses, but often this is not the case.
Discrete character data are commonly recycled from phylogenetic studies \citep[][e.g]{Foote1989-fd, Deline2018-le}.
% SG: Foote (1989) is a study of trilobite disparity from Fourier analysis of cranidia. I don't think it has anything to do with discrete data or phylogenetic studies
% MH: Also the Foote papers that used discrete data weren't taken from phylogenetic studies but collected for the specific purpose of the project -- I think this might be true for the Deline paper as well. But there are many studies out there that used matrices initially intended for phylogenetics. I can look for some if needed.
This is an efficient approach to character sampling, but it may artifactually increase disparity between phylogenetically distinct groups because phylogenetic characters are often collected to discriminate among groups \href{https://paperpile.com/c/sTGYvp/fTJ3}{(Foote 1995)}.
This needs to be considered when interpreting results, especially as synapomorphies will naturally lead to an apparent shift or increase in disparity when new clades appear.
Furthermore, many datasets are limited to subsets of anatomy that are at least implicit samples of overall anatomy, but explicit tests of this assumption have shown that different aspects of morphology can exhibit different patterns of disparity \href{https://paperpile.com/c/sTGYvp/xLdm}{(Melanie J. Hopkins 2017)}.
This effect of anatomical part on disparity patterns can be especially challenging when working on datasets where the available data has non-random missing anatomical parts, such as the absence of soft tissue in the fossil record \href{https://paperpile.com/c/sTGYvp/0y4V}{(Deline et al. 2018)}.
